---
title: "Class Grades"
format: html
---


# Learning Objectives:

- Have some fun.
- Learn about notebooks.
- What is a data frame?
  - Rows V columns
  - What is a row of data?
  - Data types
- Tour of the RStudio interface
- How to run a code chunk
- How to render a notebook
- Structure of a R command (function)


# Notebooks

- Our work will be managed in documents called Quarto notebooks.
    - This document is a notebook.
    - The weekly lecture notes are a notebook.
- A notebook combines plain text with R code to create a reproducible analysis.


# Setup

- Base R comes with some nice features, but we are going to use packages to give us additional functionality.
- We have to lead these at the beginning of the session.
- The `setup`  code chunk will always be the _first_ code chunk.
  - I will often use it to import data.
  - This code chunk loads two packages, `openxlsx` and the `tidyverse`.
    - `openxlsx`: Imports/
- Run the following code chunk to formally start your R programming journey.
  - Use the little green arrow to the right of the editor window.
  - Or use the keyboard shortcut, `Ctrl+Shift+Enter`

```{r}
#| label: setup
#| echo: false

# The first two lines of this code chunk:
# 1. Name it "setup".
# 2. Suppress all output from the code.

# Loads two packages into our current session.
library(janitor)
library(readxl)
library(tidyverse) # <- This always generates a bunch of output.

# Imports the class_grades data we used earlier.
class_grades <- read_xlsx("data/class_grades.xlsx") |> clean_names()
```

- A R function is often a verb, such as `read_xlsx()`.
- Some functions, like `library()` are not verbs. 🤷
- Immediately after a function is:
  - an "open parenthesis",
  - followed by one or more "parameters",
  - followed by a "close parenthesis".
- R code is composed of lots of "little" functions which are often chained together.
  - This is also called functional composition. For our purposes, these two terms have basically the same meaning.
  - This differs from SAS/SPSS syntax which is based around a smaller set of functions which are more powerful.


# Data

Always look at your data.

- Maybe this means opening it in Excel.
- Maybe this means looking at it in R.

Either way, look through your data.

```{r}
# Do you notice anything different about the data we are working with today?
class_grades
```

- The results of the code are embedded in the notebook.
  - When you run the code in a notebook, R creates a new file, `lab.nb.html`, which contains the original code AND the HTML containing the output.
  - In a healthcare setting, the contetns of `lab_files` folder must be carefully protected because it can contain PHI.
  - The notebook, `lab.qmd`, only contains PHI if you enter it manually into the text or code, and why on earth would you do that?
- `class_grades` is a data frame
- What does a single row of data represent?
- What kinds of data types do we have?
  - Computer Science:
    - Integers - `age`
    - Floats - `first_lab`
    - Strings - `student_name`, `grade`
    - Boolean - `TRUE`/`FALSE`; no examples here
    - Dates - `2025-09-01`; no examples here
  - Statistics:
    - Categorical
      - Nominal - `sex`; no order
      - Ordinal - `grade`; meaningful order
    - Numeric
      - Discrete - `age`
      - Continuous - `first_lab`


# Plotting

- Our first few lectures/labs are focused on plotting/graphing data.
- A plot is a graphical form of summarization.
  - From many data points to a single plot.
  - Helps us focus on the patterns, not the individual details.
- To run a single line of code within a code chunk, `Ctrl+Enter`.

How many students in the class are in each grade?

```{r}
## Look!!!
## We are using three functions to draw a plot!
## This is a form of function chaining/composition.
ggplot(data = class_grades, mapping = aes(x = grade)) +
  geom_bar() +
  ggtitle("Count of Students by Grade")
```

⬆️ Is just a plain image.

```{r}
## Look!!!
## We are using TWO (2) functions to draw a plot!
## This is a form of function chaining/composition.
library(ggiraph)
p <- ggplot(data = class_grades, mapping = aes(x = grade)) +
  geom_bar_interactive() +
  ggtitle("Count of Students by Grade")
girafe(p)
```

⬆️ This is an interactive graph.

## YOUR TURN!!!

```{r}
# Draw a bar plot showing the number of students in each lab group.

```


# Per Student Average

- Earlier in the week, we calculated per-student grades.
- Let's do that again, using R.

```{r}
# There's a lot going on here, but it isn't that different from Excel.
class_grades <-
  class_grades |>
  mutate(
    lab_grade = (first_lab + second_lab + third_lab) / 3,
    exam_grade = (first_exam + second_exam) / 2,
    final_grade = (lab_grade + exam_grade) / 2
  )
class_grades
```

⬆️ There are some differences, but this isn't that different from Excel.


# Group Averages

## Class Average

- Question: Which grade had the highest average lab grade?

```{r}
class_grades |>
  summarize(
    lab_average = mean(lab_grade),
    exam_average = mean(exam_grade),
    final_average = mean(final_grade)
  )
```

## Grade Averages

This is an example of how I structure lab questions.

- Question: Which lab group had the highest average lab grade?
- Answer: 
- Helpful Hints:
  - You can use the code we wrote above.

```{r}

```
